---
title: "GNSS vs. RTK-GNSS Positioning of Forest Inventory Plots"
author: "Florian Franz"
format:
  html:
    toc: true
    theme: lumen
---

## Indroduction

Global Navigation Satellite Systems (GNSS) are widely used for geolocating terrestrial sample plots in forest inventories. However, the accuracy of GNSS positioning varies due to factors such as the number and positions of satellites, obstacles affecting signal reception, and atmospheric conditions (Johnson & Barton, 2004). In dense forest environments, these effects are particularly pronounced, and positional errors of standard GNSS receivers can reach up to 5–10 m (McRoberts et al., 2018).

Real-Time Kinematic (RTK) GNSS correction significantly reduces these errors by using a fixed reference station or network to provide real-time differential corrections, achieving sub-meter positional accuracy (Pascual et al., 2020).

Precise plot positions are crucial in the context of remote sensing applications in forestry. Wall-to-wall datasets such as airborne laser scanning (ALS), aerial imagery, or satellite-derived canopy metrics are increasingly used to model and predict forest inventory attributes. Misregistration between field plots and remote sensing data directly increases model uncertainty (Frazer et al, 2011, McRoberts et al., 2018).

## Analysis

This analysis compares the positional accuracy of GNSS and RTK-corrected GNSS measurements for terrestrial forest inventory sample plots. By assessing the differences between both positioning methods, the potential improvement in plot coordinate accuracy achieved through RTK correction is quantified, and its implications for the co-registration of field plots with remote sensing data are evaluated.

```{r}
#| label: setup
#| include: false
proj_root <- if (file.exists('src/setup.R')) '.' else '..'
knitr::opts_knit$set(root.dir = normalizePath(proj_root))
source(file.path(proj_root, 'src/setup.R'), local = TRUE)
```

In Germany, forest management falls under the responsibility of the individual federal states. In Lower Saxony, the state forest service (Niedersächsische Landesforsten) monitors forest resources through permanent inventory plots (Betriebsinventur, BI) surveyed at 10-year intervals. The plot centers were originally positioned using standard GNSS. In the Solling region, a low mountain range in southern Lower Saxony, a subset of these plots (n = 61) was remeasured using RTK-GNSS during December 2023 and January 2024. A second remeasurement campaign covering additional plots (n = 122) was carried out from December 2025 to March 2026.

```{r}
#| label: data-reading
#| include: false
bi_plots_1 <- st_read(
  file.path(processed_data_dir, 'forest_inventory',
            'bi_center_points_2023_24.gpkg'),
  quiet = T
)

bi_plots_2 <- st_read(
  file.path(processed_data_dir, 'forest_inventory',
            'bi_center_points_2025_26.gpkg'),
  quiet = T
)

bi_plots <- st_read(
  file.path(processed_data_dir, 'forest_inventory',
            'bi_center_points_merged.gpkg'),
  quiet = T
)
```

```{r}
#| label: fig-plot-map-campaigns
#| echo: false
#| fig-cap: "Distribution of RTK-remeasured forest inventory plots in the Solling region."
bi_plots_1$campaign <- '2023/24'
bi_plots_2$campaign <- '2025/26'
bi_all <- rbind(bi_plots_1[, 'campaign'], bi_plots_2[, 'campaign'])
mapview::mapview(bi_all, zcol = 'campaign', layer.name = 'Campaign')
```

### Per-campaign details

For the first campaign, a [Trimble R2 GNSS receiver](https://utilities.trimble.com/de-de/products/trimble-r2) was used, while for the second campaign, an [Emlid Reach RS3](https://eu.store.emlid.com/de-de/products/reach-rs3) was used. Both are multiband RTK-GNSS receivers. The RTK receiver reports a solution status for each measurement. A *fix* solution indicates that the carrier-phase ambiguities were resolved, which corresponds to centimeter-level accuracy. A *float* solution means that the ambiguities were not fully resolved, leading to lower accuracy in the range of decimeters to meters. A *single* solution indicates that no RTK correction was available and the position is based on standalone GNSS, with errors typically in the range of several meters.

For the first campaign, two plots were recorded with a *single* solution. They were excluded from the further analysis. For the second campaign, six plots could not receive RTK corrections in the field. These plots were subsequently post-processed using Post-Processed Kinematic (PPK) correction and are retained in the analysis. As a result, these plots appear as fix or float solutions in the table below.

::: {layout-ncol=2}

```{r}
#| label: tbl-solution-status-1
#| echo: false
#| tbl-cap: "Campaign 1 (2023/24)"
bi_plots_1 %>%
  sf::st_drop_geometry() %>%
  dplyr::count(solution_status) %>%
  dplyr::mutate(pct = round(n / sum(n) * 100, 1)) %>%
  knitr::kable(col.names = c('Solution Status', 'n', '%'))
```

```{r}
#| label: tbl-solution-status-2
#| echo: false
#| tbl-cap: "Campaign 2 (2025/26)"
bi_plots_2 %>%
  sf::st_drop_geometry() %>%
  dplyr::count(solution_status) %>%
  dplyr::mutate(pct = round(n / sum(n) * 100, 1)) %>%
  knitr::kable(col.names = c('Solution Status', 'n', '%'))
```
:::

::: {layout-ncol=2}

```{r}
#| label: fig-solution-status-map-1
#| echo: false
#| fig-cap: "Solution status — Campaign 1 (2023/24)"
mapview::mapview(bi_plots_1, zcol = 'solution_status', layer.name = 'Solution Status')
```

```{r}
#| label: fig-solution-status-map-2
#| echo: false
#| fig-cap: "Solution status — Campaign 2 (2025/26)"
mapview::mapview(bi_plots_2, zcol = 'solution_status', layer.name = 'Solution Status')
```

:::

The center of each inventory plot is physically marked with an iron bar in the ground. However, the iron bar was not always found during the remeasurement, for example when it was located in a skid trail. In such cases, the plot center position was estimated via triangulation using nearby trees identified from stem distribution maps.

::: {layout-ncol=2}

```{r}
#| label: tbl-estimated-1
#| echo: false
#| tbl-cap: "Estimated center points — Campaign 1 (2023/24)"
bi_plots_1 %>%
  sf::st_drop_geometry() %>%
  dplyr::count(center_point_estimated) %>%
  dplyr::mutate(pct = round(n / sum(n) * 100, 1)) %>%
  knitr::kable(col.names = c('Center Point Estimated', 'n', '%'))
```

```{r}
#| label: tbl-estimated-2
#| echo: false
#| tbl-cap: "Estimated center points — Campaign 2 (2025/26)"
bi_plots_2 %>%
  sf::st_drop_geometry() %>%
  dplyr::count(center_point_estimated) %>%
  dplyr::mutate(pct = round(n / sum(n) * 100, 1)) %>%
  knitr::kable(col.names = c('Center Point Estimated', 'n', '%'))
```

:::

::: {layout-ncol=2}

```{r}
#| label: fig-estimated-map-1
#| echo: false
#| fig-cap: "Estimated center points — Campaign 1 (2023/24)"
mapview::mapview(bi_plots_1, zcol = 'center_point_estimated', layer.name = 'Estimated Center Point')
```

```{r}
#| label: fig-estimated-map-2
#| echo: false
#| fig-cap: "Estimated center points — Campaign 2 (2025/26)"
mapview::mapview(bi_plots_2, zcol = 'center_point_estimated', layer.name = 'Estimated Center Point')
```

:::

### Combined overview

Combining both campaigns results in a **total of 183 remeasured plots** covering an area of approximately 20 km² in the Solling region.

::: {layout-ncol=2}

```{r}
#| label: tbl-solution-status-combined
#| echo: false
#| tbl-cap: 'Solution status — both campaigns'
bi_plots %>%
  sf::st_drop_geometry() %>%
  dplyr::count(solution_status) %>%
  dplyr::mutate(pct = round(n / sum(n) * 100, 1)) %>%
  knitr::kable(col.names = c('Solution Status', 'n', '%'))
```

```{r}
#| label: tbl-estimated-combined
#| echo: false
#| tbl-cap: 'Estimated center points — both campaigns'
bi_plots %>%
  sf::st_drop_geometry() %>%
  dplyr::count(center_point_estimated) %>%
  dplyr::mutate(pct = round(n / sum(n) * 100, 1)) %>%
  knitr::kable(col.names = c('Center Point Estimated', 'n', '%'))
```

:::

```{r}
#| label: fig-combined-map
#| echo: false
#| fig-cap: 'Combined overview of all RTK-remeasured plots.'
bi_combined <- rbind(
  bi_plots_1[, c('campaign', 'solution_status', 'center_point_estimated')],
  bi_plots_2[, c('campaign', 'solution_status', 'center_point_estimated')]
)

mapview::mapview(bi_combined, zcol = 'campaign', layer.name = 'Campaign') +
  mapview::mapview(bi_combined, zcol = 'solution_status', layer.name = 'Solution Status', hide = TRUE) +
  mapview::mapview(bi_combined, zcol = 'center_point_estimated', layer.name = 'Estimated Center Point', hide = TRUE)
```

#### References

::: {style="font-size: 0.8em;"}
Frazer, G.W., Magnussen, S., Wulder, M.A., & Niemann, K.O. (2011). Simulated impact of sample plot size and co-registration error on the accuracy and uncertainty of LiDAR-derived estimates of forest stand biomass. Remote Sensing of Environment, 115(2), 636–649. https://doi.org/10.1016/j.rse.2010.10.008

Johnson, C.E., & Barton, C.C. (2004). Where in the world are my field plots? Using GPS effectively in environmental field studies. Frontiers in Ecology and the Environment, 2(9), 475–482. https://doi.org/10.1890/1540-9295(2004)002[0475:WITWAM]2.0.CO;2

McRoberts, R.E., Chen, Q., Walters, B.F., & Kaisershot, D.J. (2018). The effects of global positioning system receiver accuracy on airborne laser scanning-assisted estimates of aboveground biomass. Remote Sensing of Environment, 207, 42–49. https://doi.org/10.1016/j.rse.2017.10.062

Pascual, A., Guerra-Hernández, J., Cosenza, D.N., & Sandoval, V. (2020). The role of improved ground positioning and forest structural complexity when performing forest inventory using airborne laser scanning. Remote Sensing, 12(3), 413. https://doi.org/10.3390/rs12030413
:::
